Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-01 manuscript/manuscript.tex
updated ms

% We measure two types of covariances between allele frequency caused by
% selection: temporal autocovariances, and across-replicate covariances. First,
% positive temporal autocovariance in a neutral allele frequency's trajectory
% occurs when the allele becomes associated with a high or low fitness
% chromosomal background, and this association persists through the generations
% due to linkage disequilibrium. As long as the direction of selection remains
% constant, the fitness background is predictive of the direction in the neutral
% allele's frequency changes, creating positive covariance. Even though these
% magnitude of frequency changes at each site may be subtle, as they would be
% under polygenic selection, cumulatively these perturb neighboring sites in a
% predictable manner and build up temporal autocovariance which acts as a
% genome-wide signal of linked selection. Second, if evolution occurs in
% replicate populations undergoing convergent selection pressure, neutral sites
% linked to fitness backgrounds shared across replicates are expected to change
% in the same direction. This creates across-replicate covariance, which is a
% measure of the extent to which convergent selection pressures across replicate
% populations cause similar allele frequency changes. Finally, it is important to
% note that under the null model where the fitness differences between
% individuals are entirely random and non-heritable (e.g. when selection is not
% acting), both forms of covariances are expected to be zero.

@@

%\gc{TODO GIVE conversion \% var due to linked selection. }
Second, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Castro2019-uk}. This study includes two independent selection lines,
Longshanks 1 and 2 (LS1 and LS2), and an unselected control line (Ctrl).
Consequently, this selection experiment offers a useful control to test our
between-replicate covariances: we expect to see positive between-replicate
covariance in the comparison between the two Longshanks selection lines, but
not between the two pairwise comparisons between the control line and the two
Longshanks lines. We find that this is the case (gray confidence intervals in
Figure \ref{fig:figure-2} C), with the two Longshanks comparisons to the
control line not being significantly different from zero, while the comparison
between the two Longshanks line is statistically significantly different from
zero (CIs [0.013, 0.037]).

@@

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over 100kb genomic windows which we
refer to as \emph{windowed covariances} (Figure \ref{fig:figure-3} shows these
distributions pooling across all replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance of each tile will be noisy, due to sampling and genetic drift, and
the neutral distribution of the covariance is complicated due to linkage
disequilibria (which can occur over long physical distances in E\&R and
selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To address
this, we have developed a permutation-based procedure that constructs a null
distribution by randomly flipping the signs of the  allele frequency changes
per-genomic window. This destroys the systematic covariances created by linked
selection and creates a sampling distribution of the covariances spuriously
created by neutral genetic drift while preserving the complex dependencies
between adjacent loci created by linkage disequilibrium. This empirical neutral
null distribution is conservative in the sense that the variances of the
covariances are wider than expected under drift alone as they include the
effect of selection on the allele frequency change within a time-interval, just
not between time-intervals. We see (Figure \ref{fig:figure-3} A and B) that
windowed temporal covariances between close timepoints are skewed positive (a
heavy right tail), while between more distant timepoints these windowed
temporal covariances tend to shift to become more negative (a heavy left tail).
We quantified the degree to which the left and right tails are inflated
compared to the null distribution as a function of time, and see excesses in
both tails in Figure \ref{fig:figure-3} C. This finding is also robust to
sign-permuting allele frequency changes on a chromosome-level, the longest
extent that gametic linkage disequilibria can extend (Supplementary Figure
\ref{suppfig:barghi-tailprobs-seqid}). We see a striking pattern that the
windowed covariances not only decay towards zero, but in fact become negative
through time, consistent with many regions in the genome having had a reversed
fitness effect at later timepoints.

@@

We believe this is driven by the idiosyncratic nature of LD in
evolve-and-resequence populations \parencite{Nuzhdin2013-gf,Kelly2019-dc}, and
that we might find such a relationship if LD data from the evolved population
were available. Unfortunately, sequencing multiple timepoints is expensive and
the low-coverage and/or pooled-sequencing approaches common in these studies
prohibit estimating linkage disequilibrium. Future studies complete with LD
data and recombination maps would allow one to disentangle the influence of
closely linked sites from more distant sites in causing temporal
autocovariance, and allow us to better understand whether sites in high
recombination regions are free from the effects of linked selection.
Furthermore, such additional data would allow for localizing the effects of
temporal covariance, which would allow us to estimate not just the genome-wide
fraction of variation in frequency change ($G(t)$), but also the fraction of
the genome impacted by linked selection, allowing us to synthesize the temporal
approach with single-timepoint studies like that of \textcite{Elyashiv2016-vt}.

@@

% \subsection{The Empirical Neutral Null Windowed Covariance Distribution}
% \label{supp:empirical-null}

% XXX

>>>>

@@

%\gc{TODO GIVE conversion \% var due to linked selection. }
Second, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Castro2019-uk,Marchini2014-de}. This study includes two independent selection lines,
Longshanks 1 and 2 (LS1 and LS2), and an unselected control line (Ctrl).
Consequently, this selection experiment offers a useful control to test our
between-replicate covariances: we expect to see positive between-replicate
covariance in the comparison between the two Longshanks selection lines, but
not between the two pairwise comparisons between the control line and the two
Longshanks lines. We find that this is the case (gray confidence intervals in
Figure \ref{fig:figure-2} C), with the two Longshanks comparisons to the
control line not being significantly different from zero, while the comparison
between the two Longshanks line is statistically significantly different from
zero (CIs [0.013, 0.037]).

@@

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over 100kb genomic windows which we
refer to as \emph{windowed covariances} (Figure \ref{fig:figure-3} shows these
distributions pooling across all replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance of each tile will be noisy, due to sampling and genetic drift, and
the neutral distribution of the covariance is complicated due to linkage
disequilibria (which can occur over long physical distances in E\&R and
selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To address
this, we have developed a permutation-based procedure that constructs an
empirical null distribution by randomly flipping the signs of the allele
frequency changes per-genomic window. This destroys the systematic covariances
created by linked selection and creates a sampling distribution of the
covariances spuriously created by neutral genetic drift while preserving the
complex dependencies between adjacent loci created by linkage disequilibrium.
This empirical neutral null distribution is conservative in the sense that the
variances of the covariances are wider than expected under drift alone as they
include the effect of selection on the allele frequency change within a
time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that windowed temporal covariances between close
timepoints are skewed positive (a heavy right tail), while between more distant
timepoints these windowed temporal covariances tend to shift to become more
negative (a heavy left tail).  We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3} C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (Supplementary Figure \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

We believe this is driven by the idiosyncratic nature of LD in
evolve-and-resequence populations
\parencite{Nuzhdin2013-gf,Kelly2019-dc,Franssen2015-uc} \vince{put the Franssen
citation Christian suggested here}, and that we might find such a relationship
if LD data from the evolved population were available.  Unfortunately,
sequencing multiple timepoints is expensive and the low-coverage and/or
pooled-sequencing approaches common in these studies prohibit estimating
linkage disequilibrium. Future studies complete with LD data and recombination
maps would allow one to disentangle the influence of closely linked sites from
more distant sites in causing temporal autocovariance, and allow us to better
understand whether sites in high recombination regions are free from the
effects of linked selection.  Furthermore, such additional data would allow for
localizing the effects of temporal covariance, which would allow us to estimate
not just the genome-wide fraction of variation in frequency change ($G(t)$),
but also the fraction of the genome impacted by linked selection, allowing us
to synthesize the temporal approach with single-timepoint studies like that of
\textcite{Elyashiv2016-vt}.

@@

\subsection{The Empirical Neutral Null Windowed Covariance Distribution}
\label{supp:empirical-null}

To detect an excess of genomic regions with unusually high or low covariances,
we need to compare the distribution of observed windowed covariances to a null
distribution of windowed covariances that we would expect under no selection.
While we could construct a theoretic sampling distribution of the spurious
covariances created by neutral genetic drift at particular site, the unknown
linkage disequilibrium between sites would mean that this is not an adequate
null model for the distribution of windowed covariances in our data.

To address this limitation, we construct a neutral null model with the most
important feature of the neutral null: the systematic covariances between
allele frequency changes are destroyed by sign-flipping the allele frequency
changes. This is a conservative approach, since not only does selection create
temporal covariance, but it also acts to increase the magnitude of allele
frequency changes (see equation 1 of \cite{Buffalo2019-io}).

Since variants have unknown dependency structure due to LD, we sign-permute at
the window level, analogous to our block bootstrap approach.
